6.878 Final Report

### 2013-12-09 report/paper.tex
revised paper

% Note:  comment out one of the two documentclass commands, depending
% on whether you need the final or preprint version. It is most convenient
% to work in preprint mode, and switch to final only at the end.
\usepackage{float}
\usepackage{graphicx}                            % Graphics package
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage[margin=1in]{geometry}

@@

For example, the sequence $10$, $100$, $8$, $11$, $10$ is converted to \verb!S!, \verb!D!, \verb!R!, \verb!S!. Sequences were aligned and scored using Needleman-Wunsch.
In order to potentially align signals that have similar but negative expressions of each other, we attempted to give a smaller mismatch penalty between \verb!R! and \verb!D!.
We also give a very large gap penalty for large differences in time.
While sequence alignment is polynomial, we aligned on the order of $10^5$ sequences, so we modified our inital sequence alignment implementation to include gap penalty lookup as a speedup.
Alignment was run on a combination of AWS nodes and the Broad cluster.

\begin{center}
\begin{tabular}{c | ccc}
  & R & D & S\\ \hline

@@

  R &  3 & -3 & -3\\
  D & -3 & 3 & -3\\
  S & -3 & -3 & 3\\
\end{tabular}
\end{center}

@@

\subsection{Enrichment Analysis}
After identifying clusters, we wanted a way to efficiently describe each cluster, as well as to measure the effectiveness of our alignment and clustering algorithm.
To analyze the efficacy of alignment in clustering, we used Fischer's Exact test for Gene Ontology annotations.
Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.
While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions.\\

Additionally, we used \verb!DAVID! Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$. We were able to generate functional groups for several clusters.
The \verb!DAVID! clustering provided functional groups within each cluster, whereas corrected Fischer's Exact test tries to summarize an entire cluster. Since we were able to find significant groups within each cluster, but not for an entire cluster, this may indicate that our clustering is too general or that the clusters are too large.

\begin{figure}[h]
\includegraphics[height=3in]{dendrogram.png}
\caption{Dendrogram describing a cluster using our visualization tool.}
\label{fig:dendrogram}
\end{figure}

\subsection{Visualization}
In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID! gene groups by cluster.
\section{Future Goals}

We could integrate this with the SMART Genomics API to make our tool usable for all time series of gene expression.
\section{Commetary on experience}
Iris: I learned a lot on how to pull data and to compare results with the ones in a paper. I also learned a lot about

@@

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the processing for the GSE675 dataset.

>>>>

% Note:  comment out one of the two documentclass commands, depending
% on whether you need the final or preprint version. It is most convenient
% to work in preprint mode, and switch to final only at the end.
\usepackage{float}
\usepackage{graphicx}                            % Graphics package
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage[margin=1in]{geometry}
\usepackage{subcaption}

\captionsetup{compatibility=false}

@@

For example, the sequence $10$, $100$, $8$, $11$, $10$ is converted to \verb!S!, \verb!D!, \verb!R!, \verb!S!. Sequences were aligned and scored using Needleman-Wunsch.
In order to potentially align signals that have similar but negative expressions of each other, we attempted to give a smaller mismatch penalty between \verb!R! and \verb!D!. We also tried stricter versions of the scoring matrix to allow softened mismatch (between steady and anything else) or no mismatches (Table \ref({tab:scoring}).
We also varied the gap penalty depending on how far apart the time points were. For very far apart signals, we used a larger penalty.
While sequence alignment is polynomial, we aligned on the order of $10^5$ sequences, so we modified our inital sequence alignment implementation to include gap penalty lookup as a speedup.
Alignment was run on a combination of AWS nodes and the Broad cluster.

\begin{table}
\caption{Some sample scoring matrices alignment: least strict (allowing matches between drops and rises in signal for potention downregulation effects), more strict (giving mismatches between rises and drops), most strict (penalizing mismatches). The gap penalty depends on the time offset between two time points, increasing with time differences.}
\label{tab:scoring}
\begin{center}
\begin{tabular}{c | ccc}
  & R & D & S\\ \hline

@@

  R &  3 & -3 & -3\\
  D & -3 & 3 & -3\\
  S & -3 & -3 & 3\\
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Enrichment Analysis}
After identifying clusters, we wanted a way to efficiently describe each cluster, as well as to measure the effectiveness of our alignment and clustering algorithm.
To analyze the efficacy of alignment in clustering, we used Fischer's Exact test for Gene Ontology annotations.
Gerald Quon provided us with an \verb!R! script and Gene Ontology data for annotation testing of our clusters.
While, visually, the clusters showed alignment, the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure. However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig. \ref({fig:enrichment}). It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\

In order to provide a summarizing description of learned clusters we wanted to provide a word cloud, we initially pulled gene information from the \verb!OMIM! database. However, the text was too general even after multiple round of filtering and the generated word clouds were too non-descriptive.

We then used \verb!DAVID! Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$. We were able to generate functional groups for several clusters (Fig. \ref({fig:enrichment}).
The \verb!DAVID! clustering provided functional groups within each cluster, whereas corrected Fischer's Exact test tries to summarize an entire cluster. Since we were able to find significant groups within each cluster, but not for an entire cluster, this may indicate that our clustering is too general or that the clusters are too large.

\begin{figure}[h]

\end{figure}

\begin{figure}[H]
  \centering
  \begin{subfigure}[b]{0.5\textwidth}
    \includegraphics[width=.9\textwidth]{dendrogram.png}
    \label{fig:dendrogram}
  \end{subfigure}%
  ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
    %(or a blank line to force the subfigure onto a new line)
  \begin{subfigure}[b]{0.5\textwidth}
    \includegraphics[width=.9\textwidth]{wordcloud.png}
    \label{fig:wordcloud}
  \end{subfigure}%
  \label{fig:enrichment}

  \begin{subfigure}[b]{0.8\textwidth}
    \includegraphics[width=.7\textwidth]{pvalue_performace.png}
    \label{fig:pvalue}
  \end{subfigure}%
  \label{fig:enrichment}
  \caption{A) Dendrogram describing a cluster using our visualization tool. B) A word cloud generated from the clusters in the dendrogram. C) To give some performance metric, we took the ratio of the p-values for enrichment analysis for our technique over using plain k-means for the GSE675 dataset.}
\end{figure}

\subsection{Visualization}
In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID! gene groups by cluster (Fig. \ref({fig:enrichment}).
\section{Future Goals}
We would re-do the pre-processing step or attempt to get the pre-processed dataset before reanalyzing our data.

We could integrate this with the SMART Genomics API to make our tool usable for all time series of gene expression.

\section{Comparison with original proposal} 
We managed to complete everything that we proposed, including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections). We also provided a visualization tool as part of our combined class project.

\section{Commetary on experience}
Iris: I learned a lot on how to pull data and to compare results with the ones in a paper. I also learned a lot about pre-processing data and steps to take when results don't match those of the paper.

@@

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
